Transfer learning between Sentinel-1 acquisition modes enhances \\ the few-shot segmentation of natural oil slicks in the Arctic

### 2025-04-06 latex/manuscript.tex
conclusion structure

% The paper headers
\markboth{\LaTeX\ PDF, IEEE GEOSCIENCE AND REMOTE SENSING LETTERS,~Vol.~XX, No.~Y, <Month>~2025}%
{Shell \MakeLowercase{\textit{et al.}}: A Sample Article Using IEEEtran.cls for IEEE Journals}

@@

Previous works have attempted to make use of the damping ratio for automatizing marine oil detection, with varying results. They face a common challenge which has been an ongoing topic of research for over three decades: 
the presence of oil look-alikes including ocean eddies, biogenic slicks and newly-formed ice \cite{johanssonCanMineralOil2020,hovlandSlickDetectionSAR1994,alpersOilsSurfactants2004,alpersOilSpillDetection2017}. 
When trained with a large number of labeled examples, deep learning models, currently the benchmark in computer vision, can capture complex shape- and texture-related features for image classification or segmentation tasks 
\cite{goodfellowDeepLearning2016}. When trained with a limited sample size, however, these models struggle to extract more abstract, higher-level features
\cite{bengioDeepLearningRepresentations2012}. The issue of limited data is exacerbated in the case of natural slicks because of limited visibility. Not only will slicks only last a few hours on the surface 
\cite{jatiaultMonitoringNaturalOil2017,daneshgaraslHindcastModelingOil2017,oreillyDistributionMagnitudeVariability2022}, their visibility also depend on a specific window of surface wind conditions, outside of which the sea 
either has too many or too few capillary waves for a discriminative damping ratio \cite{quigleyInvestigationDampingRatio2023,sausDetectionDelineationProduced2021,gadeImagingBiogenicAnthropogenic1998}. 
\IEEEpubidadjcol  
Slick events are also highly intermittent. Seepage rates vary across both monthly and yearly timescales, and most active seeps have an occurrence rate below 10\% 
\cite{jatiaultNaturalOilSeep2024,oreillyDistributionMagnitudeVariability2022}. Some seeps remain dormant for months, or even years, before emitting oil again. In two distinct regions, studies found that 
only half of all seeps show activity at a given time\cite{jatiaultMonitoringNaturalOil2017,garcia-pinedaRemotesensingEvaluationGeophysical2010}, and dormant seeps' occurrence might be underestimated, 
considering that they are less likely to be discovered.

In this paper, we assess the potential of transfer learning for enhancing automatic mapping of natural oil slicks. Transfer learning, or deep transfer learning in the case of deep learning, aims at 
transferring valuable and generalizable features from one model to another, where contexts are similar but the target distributions differ \cite{goodfellowDeepLearning2016}. 
We pretrained two image segmentation models using oil slicks mapped in Sentinel-1 images over the North Sea. These models were then fine-tuned on two known seeps in the Arctic, for which we have only 
limited observations. In the Arctic, Sentinel-1 scenes are mostly acquired in the extra wide (EW) mode at a 25-m pixel spacing with dual HH/HV polarization. However, elsewhere, oil mapping with Sentinel-1 more 
commonly uses the interferometric wide (IW) mode at a 10-m pixel spacing with VV/VH polarization, as it is the standard mode in oceans outside of polar regions.
%Over the Arctic, Sentinel-1 scenes are mostly acquired in the 25-m resolution extra-wide (EW) mode with a dual HH/HV polarization, whereas oil slick mapping with Sentinel-1 more often use the 
%interferometric-wide (IW) mode, of 10-m pixel resolution with a dual VV/VH polarization. 
Little work, if any, has addressed transfer learning between SAR acquisition modes, and we believe natural oil slicks make a prime example of a real-life application of deep-learning mapping with few labeled data.

@@

Using all 463 images from the $\mathit{D}_S$, we pretrained models for a maximum of 100 epochs with validation-loss early-stopping as a means of limiting overfitting \cite{wangGeneralizingFewExamples2020}. 
We then generated 8 bootstrapped sets for each value in the set \( N = \{ 2, 8, 32, 128\} \), where \( N \) is the number of training images from the $\mathit{D}_T$. No bootstrapping was done at \( N = 128 \) 
as the entire training split was in use. We compared \textit{base models}, which used default backbone weights, and \textit{fine-tuned models}, which were instead initialized with full weights from the model 
pretrained on the $\mathit{D}_S$. These weights are left unfrozen in the $\mathit{D}_T$ fine-tuning stage \cite{yosinskiHowTransferableAre2014}. We calculated the Dice score for a total of 128 models on the test split, 
considering only the oil slick class to ignore the trivial classification of water. The Dice score, closely related to the F1, is calculated as:
\begin{align}
    \text{Dice} = \frac{2 \cdot TP}{2 \cdot TP + FP + FN}
\end{align}
Bootstrapped testing helps assess model consistency with small and high-variance test sets \cite{mollersenAccountingMultiplicityMachine2024}. In Fig.\,\ref{figure2} and Table\,\ref{table2},
we provide a 95\% statistical interval for the test-set Dice score, following the bootstrapping method described in
\cite{burkovMachineLearningEngineering2020}.

\section{Results and Discussion} 
Fig. \ref{figure2} illustrates the change in Dice score as the number of training images increases. We show clear evidence that both the convolutional neural network (Fig. \ref{fig2-1}) 
and the vision transformer (Fig. \ref{fig2-2}) benefit from transfer learning. This result is more evident in the case of DeepLabv3, in which $\mathit{D}_S$ pretraining increased the Dice score at all values of \( N \).
While SamLoRA also benefits, the effect is rapidly reduced as the number of training samples grow, and its effect is indistinguishable with values of \( N > 8 \). Despite this, transfer learning has an overall positive effect, 
particularly with fewer training images. Until \( N > 8 \), SamLoRA outperforms DeepLabv3. This is consistent with larger models being more efficient at fine-tuning on small number of samples \cite{zhaiScalingVisionTransformers2022}. 
DeepLabv3's Dice scores grow steadily with an increasing number of training images, and we can expect this trend to continue beyond the limits of our sample size (Fig. \ref{fig2-1}). In contrast, it remains unclear whether SamLoRA 
would benefit from transfer learning with larger training datasets (Fig. \ref{fig2-2}).
\begin{table}[t]
    \caption{Average Dice and Change Observed with Transfer Learning, Based on the Spatial Alignment Method\label{table2}}
    \centering
    \begin{tabular}{c>{\centering\arraybackslash}m{3cm}>{\centering\arraybackslash}m{2.6cm}}
    \toprule
    \textbf{Tile size} & \textbf{25\,m downsampling} & \textbf{10\,m upsampling} \\
    \midrule
    \textbf{256x256} & $0.49 (+0.13 \pm 0.13)$ & $0.49 (+0.05 \pm 0.04)$ \\
    \textbf{512x512} & $0.42 (+0.11 \pm 0.28)$ & $0.51 (+0.03 \pm 0.05)$ \\
    \bottomrule
    \end{tabular}
\end{table}

After comparing model architectures, we investigated the effect of transfer learning depending on the resampling method used to geographically align images from two Sentinel-1 acquisition modes. Table\,\ref{table2} 
compares these methods, presenting the average Dice score accompanied by a 95\% statistical interval for the change in Dice resulting from transfer learning. In most cases, transfer learning improves model performance, 
but its effect is strongest—yet highly variable—when downsampling images to 25\,m, meaning that we align the $\mathit{D}_S$ IW images with the $\mathit{D}_T$ EW images. Similar work by 
\cite{bianchiLargeScaleDetectionCategorization2020a} found that fine-tuning improved deep learning segmentation of oil spills. In that study, however, the researchers downsampled the resolution of training images to create a 
synthetic pretraining dataset.
Moreover, tile size has no apparent effect on transfer learning. This is an interesting finding, because large geophysical processes, such as oil slicks, often need to be split into smaller tiles for processing. This 
typically results in performance loss and slower computing.

\begin{figure*}[!t]
    \centering
    \includegraphics[width=6.02in]{figures/preds4x3_v2_scale_lettering.png}
    \caption{Top-1 model performance through examples. \quad Top row: An easier case, with some ocean eddies. \quad Middle: A more challenging example with 
    various look-alike features. \quad Bottom: Arguably one of the most challenging cases. A small, low-contrast slick is missed by models in the center right of the image.}
    \label{figure3}
\end{figure*}
Fig.\,\ref{figure3} displays the segmentation predictions from the top-performing model (Top-1). We compare predictions from the base (Fig.\,\ref{figure3}b) and the fine-tuned (Fig.\,\ref{figure3}c)
versions, showing how the model benefited from transfer learning across different examples. The examples are taken from the test set, with a 25\,m resolution and predicting 
with a tile width of 512 pixels, the most challenging format evaluated (Table\,\ref{table2}). First, we see a typical arc-shaped slick with a clear SAR damping ratio (top, Fig.\,\ref{figure3}a). 
Both the base and the fine-tuned model delineate the slick while correctly leaving out ocean eddies and a low-wind area on top of the image. The second row presents a common appearance of natural slicks in an SAR image, 
where several dark features may trigger false detections (middle, Fig. \ref{figure3}a). Nonetheless, both models detect the two slicks (middle, Fig. \ref{figure3}d) and transfer learning 
help reduce false detections, despite a somewhat undervalued Dice score (middle, Fig.\,\ref{figure3}c). The last row (bottom, Fig.\,\ref{figure3}a) exemplifies one of the most challenging cases for SAR slick mapping. 
If two low-contrast slicks can be seen among look-alike features, only the main slick (bottom, Fig. \ref{figure3}d) is detected by the fine-tuned model (bottom, Fig. \ref{figure3}c), illustrating the challenge of 
harsh mapping conditions.

\section{Conclusion}
Applications of transfer learning must be carefully evaluated. In many cases, its use can be inefficient, even detrimental, if the contexts are too different \cite{mensinkFactorsInfluenceTransfer2022}.
In this study, we made use of labeled slicks from a different area and imaged in a different SAR acquisition mode. We show that transfer learning in the form of fine-tuning is effective for the few-shot SAR segmentation
of natural oil slicks, improving model performance (average Dice increase of 0.03--0.13) despite accessing only a few examples in the target domain.

Transferability between IW and EW modes is especially important in the northern, where the absence of large spills--which oil segmentation models typically rely on for training data-- 
limits the availability of training data. In contrast, seeps in the region emit slicks that are typically thin, with low oil concentrations and reduced backscatter damping.
This poses a challenge for detection, which partly explains why these remote seeps are still under-investigated today. Our results show that both deep learning models benefited from transferring pretrained model weights, 
with DeepLabv3 performing best overall, though SAM achieved top accuracy with fewer training images. Our method, validated on two documented seepage sites, offers promising perspectives for
identifying and studying additional, as yet undiscovered remote seeps.

>>>>

% The paper headers
% \markboth{\LaTeX\ PDF, IEEE GEOSCIENCE AND REMOTE SENSING LETTERS,~Vol.~XX, No.~Y, <Month>~2025}%
\markboth{Vadnais \MakeLowercase{\textit{et al.}}: EarthArXiv preprint, submittedApril 2025}
{Shell \MakeLowercase{\textit{Vadnais et al.}}: A Sample Article Using IEEEtran.cls for IEEE Journals}

@@

Previous works have attempted to make use of the damping ratio for automatizing marine oil detection, with varying results. They face a common challenge which has been an ongoing topic of research for over three decades: 
the presence of oil look-alikes including ocean eddies, biogenic slicks and newly-formed ice \cite{johanssonCanMineralOil2020,hovlandSlickDetectionSAR1994,alpersOilsSurfactants2004,alpersOilSpillDetection2017}. 
When trained with a large number of labeled examples, deep learning models, currently the benchmark in computer vision, can capture complex shape- and texture-related features for image classification or segmentation tasks 
\cite{goodfellowDeepLearning2016}. When trained with a limited sample size, however, these models struggle to extract more abstract, higher-level features
\cite{bengioDeepLearningRepresentations2012}. The issue of limited data is exacerbated in the case of oil slicks because of limited visibility. Not only will slicks only last a few hours on the surface 
\cite{jatiaultMonitoringNaturalOil2017,daneshgaraslHindcastModelingOil2017,oreillyDistributionMagnitudeVariability2022}, their visibility also depend on a specific window of surface wind conditions, outside of which the sea 
either has too many or too few capillary waves for a discriminative damping ratio \cite{quigleyInvestigationDampingRatio2023,sausDetectionDelineationProduced2021,gadeImagingBiogenicAnthropogenic1998}. 
\IEEEpubidadjcol  
Slick events are also highly intermittent. Seepage rates vary across both monthly and yearly timescales, and most active seeps have an occurrence rate below 10\% 
\cite{jatiaultNaturalOilSeep2024,oreillyDistributionMagnitudeVariability2022}. Some seeps remain dormant for months, or even years, before emitting oil again. In two distinct regions, studies found that 
only half of all seeps show activity at a given time\cite{jatiaultMonitoringNaturalOil2017,garcia-pinedaRemotesensingEvaluationGeophysical2010}, and dormant seeps' occurrence might be underestimated, 
considering that they are less likely to be discovered.

In this paper, we assess the potential of transfer learning for enhancing automatic mapping of natural oil slicks. Transfer learning, or deep transfer learning in the case of deep learning, aims at 
transferring valuable and generalizable features from one model to another, where contexts are similar but the target distributions differ \cite{goodfellowDeepLearning2016}. 
We pretrained two image segmentation models using oil slicks mapped in Sentinel-1 images over the North Sea. These models were then fine-tuned on two known seeps in the Arctic, for which we have only 
limited observations. In the Arctic, Sentinel-1 scenes are mostly acquired in the extra wide (EW) mode at a 25-m pixel spacing with dual HH/HV polarization. However, elsewhere, oil mapping with Sentinel-1 more 
commonly uses the interferometric wide (IW) mode at a 10-m pixel spacing with VV/VH polarization, as it is the standard mode in oceans outside of polar regions.
Little work, if any, has addressed transfer learning between SAR acquisition modes, and we believe natural oil slicks make a prime example of a real-life application of deep-learning mapping with few labeled data.

@@

Using all 463 images from the $\mathit{D}_S$, we pretrained models for a maximum of 100 epochs with validation-loss early-stopping as a means of limiting overfitting \cite{wangGeneralizingFewExamples2020}. 
We then generated 8 bootstrapped sets for each value in the set \( N = \{ 2, 8, 32, 128\} \), where \( N \) is the number of training images from the $\mathit{D}_T$. No bootstrapping was done at \( N = 128 \) 
as the entire training split was in use. We compared \textit{base models}, which used default backbone weights, and \textit{fine-tuned models}, which were instead initialized with full weights from the model 
pretrained on the $\mathit{D}_S$. These weights are left unfrozen in the $\mathit{D}_T$ fine-tuning stage \cite{yosinskiHowTransferableAre2014}. We calculated the Dice score for a total of 128 models on the test split, 
considering only the oil slick class to ignore the trivial classification of water. The Dice score, closely related to the F1, is calculated as:
\begin{align}
    \text{Dice} = \frac{2 \cdot TP}{2 \cdot TP + FP + FN}
\end{align}
Bootstrapped testing helps assess model consistency with small and high-variance test sets \cite{mollersenAccountingMultiplicityMachine2024}. In Fig. \ref{figure2} and Table \ref{table2},
we provide a 95\% statistical interval for the test-set Dice score, following the bootstrapping method described in
\cite{burkovMachineLearningEngineering2020}.

\section{Results and Discussion} 
Fig. \ref{figure2} illustrates the change in Dice score as the number of training images increases. We show clear evidence that both the convolutional neural network (Fig. \ref{fig2-1}) 
and the vision transformer (Fig. \ref{fig2-2}) benefit from transfer learning. This result is more evident in the case of DeepLabv3, in which $\mathit{D}_S$ pretraining increased the Dice score at all values of \( N \).
While SamLoRA also benefits, the effect is rapidly reduced as the number of training samples grow, and its effect is indistinguishable with values of \( N > 8 \). Despite this, transfer learning has an overall positive effect, 
particularly with fewer training images. Until \( N > 8 \), SamLoRA outperforms DeepLabv3. This is consistent with larger models being more efficient at fine-tuning on a small number of samples \cite{zhaiScalingVisionTransformers2022}. 
DeepLabv3's Dice scores grow steadily with an increasing number of training images, and we can expect this trend to continue beyond the limits of our sample size (Fig. \ref{fig2-1}). In contrast, it remains unclear whether SamLoRA 
would benefit from transfer learning with larger training datasets (Fig. \ref{fig2-2}).
\begin{table}[t]
    \caption{Average Dice and Change Observed with Transfer Learning, Based on the Spatial Alignment Method\label{table2}}
    \centering
    \begin{tabular}{c>{\centering\arraybackslash}m{3cm}>{\centering\arraybackslash}m{2.6cm}}
    \toprule
    \textbf{Tile size} & \textbf{25\,m downsampling} & \textbf{10\,m upsampling} \\
    \midrule
    \textbf{256x256} & $0.49 (+0.13 \pm 0.13)$ & $0.49 (+0.05 \pm 0.04)$ \\
    \textbf{512x512} & $0.42 (+0.11 \pm 0.28)$ & $0.51 (+0.03 \pm 0.05)$ \\
    \bottomrule
    \end{tabular}
\end{table}

After comparing model architectures, we investigated the effect of transfer learning depending on the resampling method used to geographically align images from two Sentinel-1 acquisition modes. Table \ref{table2} 
compares these methods, presenting the average Dice score accompanied by a 95\% statistical interval for the change in Dice resulting from transfer learning. In most cases, transfer learning improves model performance, 
but its effect is strongest—yet highly variable—when downsampling images to 25\,m, meaning that we align the $\mathit{D}_S$ IW images with the $\mathit{D}_T$ EW images. Similar work by 
\cite{bianchiLargeScaleDetectionCategorization2020a} found that fine-tuning improved deep learning segmentation of oil spills. In that study, however, the researchers downsampled the resolution of training images to create a 
synthetic pretraining dataset.
Moreover, tile size has no apparent effect on transfer learning. This is an interesting finding, because large geophysical processes, such as oil slicks, often need to be split into smaller tiles for processing. This 
typically results in performance loss and slower computing.

\begin{figure*}[!t]
    \centering
    \includegraphics[width=6.02in]{figures/preds4x3_v2_scale_lettering.png}
    \caption{Top-1 model performance through examples. \quad Top row: An easier case, with some ocean eddies. \quad Middle: A more challenging example with 
    various look-alike features. \quad Bottom: Arguably one of the most challenging cases. A small, low-contrast slick is missed by models in the center right of the image.}
    \label{figure3}
\end{figure*}
Fig. \ref{figure3} displays the segmentation predictions from the top-performing model (Top-1). We compare predictions from the base (Fig. \ref{figure3}b) and the fine-tuned (Fig. \ref{figure3}c)
versions, showing how the model benefited from transfer learning across different examples. The examples are taken from the test set, with a 25\,m resolution and predicting 
with a tile width of 512 pixels, the most challenging format evaluated (Table \ref{table2}). First, we see a typical arc-shaped slick with a clear SAR damping ratio (top, Fig. \ref{figure3}a). 
Both the base and the fine-tuned model delineate the slick, while correctly leaving out ocean eddies and a low-wind area on top of the image. The second row presents a common appearance of natural slicks in an SAR image, 
where several dark features may trigger false detections (middle, Fig. \ref{figure3}a). Nonetheless, both models detect the two slicks (middle, Fig. \ref{figure3}d) and transfer learning 
help reduce false detections, despite a somewhat undervalued Dice score (middle, Fig. \ref{figure3}c). The last row (bottom, Fig. \ref{figure3}a) exemplifies one of the most challenging cases for SAR slick mapping. 
If two low-contrast slicks can be seen among look-alike features, only the main slick (bottom, Fig. \ref{figure3}d) is detected by the fine-tuned model (bottom, Fig. \ref{figure3}c), illustrating the challenge of 
harsh mapping conditions.

\section{Conclusion}
Applications of transfer learning must be carefully evaluated. In many cases, its use can be inefficient, even detrimental, if the contexts are too different \cite{mensinkFactorsInfluenceTransfer2022}.
In this study, we made use of labeled slicks from a different area and imaged in a different SAR acquisition mode. We show that transfer learning in the form of fine-tuning is effective for the few-shot SAR segmentation
of natural oil slicks, improving model performance (average Dice increase of 0.03--0.13) despite accessing only a few examples in the target domain. 
Indeed, both deep learning models benefited from transferring pretrained model weights, 
with DeepLabv3 performing best overall, though SAM achieved top accuracy with fewer training images.

Transferability between IW and EW modes is especially important in the Arctic. In this region, the absence of large offshore spills--which oil segmentation models typically
rely on for training--limits the availability of training data. In contrast, Arctic seeps emit slicks that are generally thinner, with low oil concentrations and reduced backscatter damping.
This poses a challenge for detection, which partly explains why these remote seeps are still under-investigated today. Our method, validated on two documented seepage sites, offers promising perspectives for
identifying and studying additional, as yet undiscovered remote seeps.
